---
title: "Smart data Masking Function Requirements/Possibilities"
author: "Carlos Poses"
format: 
  html:
    toc: true
    number-sections: true
    toc-expand: true
    toc-location: body
editor: source
---

## Problem

-   In tables that display the counts of categorical variables, we want to mask values for categories whose count is less than 5 (or some other threshold $T$).
-   To implement this masking, we need to make sure the value of the masked value can't be computed from the other values and the totals.


I'll call 'naive masking' simply changing values less than a threshold by a string 'Naive Masking'.

In this doc: First, I explain when naive masking does not work. This results in some rules that can (probably) be implemented in a function that checks every table to check all values are properly masked. I also discuss when naive masking works, but possible solutions are small so that naive masking may still be considered ineffective. Second, I discuss what to do when naive masking does not working but we still want to give as much info as possible (and the category totals will be given).

## When naive masking does not work.

Simply put, naive masking does not work whenever the masked values (the unknowns) are a function of known values (totals and other values). Equivalently (I think), whenever the solution to the unknown values can be worked out using a equation with one unique solution.

### Example

```{r}
cat_value <- c(3, 97, 100)
total <- sum(cat_value)
total
```

Masking the value 3 (e.g., substituting it by '<= 5') is not enough if we know the other values and the total: 200-97-100 = 3.

### Expressing the problem using an equation

For each category, the problem can be represented with the equation:

$$ x + y + ... + z = N$$

where $x,y,...,z$ are known integers representing category counts and $N$ is the total.

Say we have a thresold ($T$, e.g. 5), and we mask values lower than this threshold. We denote masked values with superscript M (e.g., $x^M$).

Naive masking won't work whenever the masked values can be expressed as an equation with one unique solution. More specifically, I came up with 3 scenarios when naive masking won't work:

1.  **Masking one value without masking the total (or another value) is equivalent to not masking.**

$$ x + y  + ... + z = N$$ 

If we mask one value ($x^M$), this turns into an linear equation with one unknown and thus an unique solution.

$$ x^M = N - y - ... - z $$ 

(This is the case represented by the first example)

2. **Masking multiple values is equivalent to not masking in two scenarios** 

In general an equation with two or more unknowns has no solution. But given the restrictions (all values are positive integers), there are two special cases where there's only one possible solution:

2.1. **When all original masked values are one, because one can still work it out from the total.**

For instance, we mask two values. 

$$ x^M + y^M  + ... + z = N$$ 

If $x^M = y^M = 1$ then

$$ x^M + y^M = N - ... - z == 2 $$ 

Even if you don't know what the original $x^M$ and $y^M$ values are, if you know \$ N - ... - z $ and the number of masked values (\# M), the only solution is that all values are 1.
This is general always the case when the result of substracting unmasked values from the total is equal to the number of masekd values.

2.1. **When all masked values are equal to the highest possible value of the masked value**

Say threshold is less or equal than 5.

$$ x^M + y^M  + ... + z = N$$ 

If $x^M = y^M = 5$ then

$$ x^M + y^M = N - ... - z == 10 $$

Again there's only one solution for the individual values of $x^M$ and $y^M$. So this is equivalent to masking.


### Conclusion

Naive masking won't work if you can specify the counts as an equation, and it is possible to work out the unknowns (masked values), i.e., the equation has unique solutions. 

Possible solution: Try to solve the system of equations with some R library (or base R) and proper restrictions (positive integers). If there's unique solution to the there's no masking. 

NB: I am not 100% sure there're more than the 3 scenarios where masking will work, but I think framing the problem as 'is there an unique solution to this equation, where the masked values are the unknown?' should cover every possibly scenario. As long as we find a way to always answer this question (given restrictions on possible solutions), it should be fine.


## Naive masking works in a strict sense, but we give away too much info. 

### Problem

- Typically when we mask, we say a value is less than 5. What we want is to give no other information than that value is an integer in the interval [1,4]. For all categories, the values 1, 2, 3 and 4.
- Sometimes when applying naive masking, it's impossible to know the exact value, but it is still possible to exclude some values from the interval, e.g, you know the value is either 1 or 2.

### Example

```{r}
cat_value <- c(2, 1, 97, 100) 
total <- sum(cat_value)
total
```

If our threshold is 5, we would mask the values 2 (let's call it $x^M$) and 1 (let's call it $y^M$). But then we know

\$200 - 97 - 100 = 3 \$

or

$$ N - z - w = x^M + y^M$$ 

Even though we want to say that the value for both x and y is $<5$ (i.e., between 1 and 4), in this example you can tell that for x (or y) it is either 2 or 1. It's impossible to tell the exact value, because there's no unique solution. But we know the value is neither 3 nor 4.

### A trickier example

```{r}
cat_value <- c(2, 4, 92, 100) 
total <- sum(cat_value)
total
```

$198 - 192 = 6$

We would mask two values, and we know their sum is equal to 6. That makes it impossible that either value is 1.

### An example that works

```{r}
# we would mask 2 and 4, let's call them x and y
total - 100 - 97
```
You know either x or y is not 5 (because both can't be 5. Actually both can't be 4, or one 5 and another 4, etc). But you are saying nothing about the values: x could be 5, 4, 3, 2 or 1. Same for y.

You reveal some info: both masked are not close to the threshold. But you can't tell which one, and all values below the threshold are possible for both counts. 
This is the info you are giving: *at least one of the values is not equal to 4 (or equal to T - 1 in general)*. But you can't tell for *any specific value* whether it is exactly equal to 5 or not. 

### Formalizing it

When the original value is masked 'naively', we want to make sure all values in the interval are theoretically possible, given the rest of the provided information (counts for other categories and totals). This depends exactly on the other information: the remaining value after substracting the sum of the other category counts to the total. It turns out this is the case whenever the following is true:

$$ N - w - ... z \geq T + \# M - 2 $$

where T is the thresold (i.e. 5 in less than 5), $ \# M $ is the number of 'Naively masked' variables, N is the total and w,...,z are the counts for other variables.

(There's an upper threshold as well, but I don't discuss it because it's a case where the equation would have an unique solution).

Weird formula I guess, but easy (?) to see with an example. If there're two unknowns, the formula becomes:

$$ N - w - ... z \geq T $$
So you have masked two variables (x and y) and the threshold is 5. If you substract the values of the unmasked categories from the total, you get 5. Now x can be any value between 1 and 4 (and y too). If you would get 5, you would know x can't be 5 (because y would need to be 0).

If you have 3 masked variables, then the number is 6 (5 + 3 - 2). It is always possible, for all variables, to be either 1, 2, 3 and 4.

In words: if the difference between the totals and the unmasked categories is higher than the threshold + the number of masked values - 1, then:

- For a specific masked value, nobody can tell if its value is 4 or smaller than 4.
- But in many cases (e.g., when the exact equality holds), you know at least one value is smaller than five. You just don't know which one. Maybe this is enough.

#### Example

```{r}
cat_value <- c(2, 4, 97, 100)
total <- sum(cat_value)
total
```


#### Potential problem 

This strategy might seem ok, but it scales a bit badly. Say that, naively, you would mask 6 values.
```{r}
cat_value <- c(2,1,1,2,3, 4, 97, 100)
total <- sum(cat_value)
total
total - 100 - 97
```

Here you know that the sum of the 6 masked values is 13. All values are 'well masked', i.e., you can't tell for any specific value whether it's 5,4,3,2, or 1. But you are giving away the info that most values are not 5, or 4, or 3. It can be a bit problematic in some edge cases (e.g. there are 10 masked values that add up to 14, you know many values are 1).

If this is enough, one can just check this rule is satisfied.


## Possible solutions

### Problem recap

1. If you can formulate the count as an equation of the form:
$$ x^M + y^M  + ... + z = N$$ 

with one unique solution for each unknown, then you are not properly masking.
(Important: there can be as many masked variables (superscript M) and as many unmasked variables (no superscripts) as you want in the previous equation).

2. If however there's not unique solution and you have this:

$$ N - w - ... z \geq T + \# M - 2 $$ 
then you can't tell if each value is 4, 3, 2 or 1. But the closer the lhs is to the rhs, the more info you have about the possible values (i.e., you know some values are much smaller).

### Actual implementation

Write an algorithm that checks that 1 and 2 are satisfied.

- If TRUE, table is fine. Naive masking is enough (or there's no masking).
- If FALSE, table is not fine. We need to do something else.

### Possible solutions

If naively you would mask only one value, one solution is to interval-mask an additional value (e.g., the second smallest value). It's easier to reason with an example:


```{r}
cat_value <- c(3, 97, 100)
total <- sum(cat_value)
total
```

Here, naive masking is not enough. If you mask 3, you know the value from the totals. A possible solution is to 'interval-mask', e.g., 

```{r}
c('1-4', '99-95', 100)
```
This solution will work always for Scenario 1. But it doesn't work too well for Scenarios 2.1 and 2.2, I think. Imagine you have this pattern
```{r}
c(1,1,98,100)
```
This masking won't be enough, because of scenario 2.1: If you know the total, there's only one value the masked values can take.
```{r}
c('<5', '<5', 98, 100)
```
Now we apply interval-masking.
```{r}
c('<5', '<5', '95-100', 100)
```





$ x^M + z + w = N$

First, add up $x^M + z + w = N$.






In addition to the originally naively masked value, mask another value in an interval form.

Easier solution:

You can transform variables until conditions hold.

- Instead of giving information about N, you can give it in a interval: [Nmin, Nmax]

$$ N - z - 1 $$ 


you also have this

$$ N - w - ... z \geq x^M + ... + y^M $$



